---
title: 'R Project  : Exploratory Data Analysis in R'
Name: Suchita Negi
Email: snegi@scu.edu
---

## **Introduction**: Exploratory Analysis of IBM Employee Atrrition .
In EDA we visualize the factors that lead to employee attrition and explore important factors that affect the attrition in the company.

**Dataset selection: ** As my interest leads towards People Analytics, I have selected the dataset from Kaggle that can help me undersatand HR basics for analysis using R.
The data set was downloaded from Kaggle dataset:
https://www.kaggle.com/pavansubhasht/ibm-hr-analytics-attrition-dataset

```{r}
## Installing packages required for data analysis
if (!require(dplyr))
  install.packages("dplyr")
if (!require(ggplot2))
  install.packages("ggplot2")
if (!require(gridExtra))
  install.packages("gridExtra")

## Loading required libraries
library(dplyr)
library(ggplot2)
library(gridExtra)
```


## Initial Analysis

**Loading the data set in R : **
Reading the csv file located at the local directory.
The dataset selected ws clean and therefore didnot require additional cleaning.
```{r}

setwd("D:/Academic/SCU_Q1/R_Programming/R_Studio")
#Loading the dataset in R
hr_data <- read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv")
#checking the data if loaded correctly
head(hr_data)

```

**Analyse the structutre of the data : **
The dataframe consists of 1470 rows and 35 columns.
The data set have various numerical and categorical data.
```{r}
str(hr_data)
```

**Analyse the data quality : **
Analysing the data quality by summarizing dataset to observe min, max , mean , median of differnt columns.
```{r}
#summarize the data columns
summary(hr_data)
```

## Data Analysis for Attrition 

Dataset is analysed by observing different factors affecting the attrition in the company.

**Libraries used: ** "dplyr" library to manipulate dataset and subset/filter data and group gender.
                      "ggplot2" library to plot bar graph for attrition in each gender.
                      "gridextra" library to display multiple plots in grid format.

**Atrrition count with respect to Gender : **
Blue Bar: Male
Pink Bar: Female
"Yes":Employee Attrition
"No": Employee Retained
Observation:
We observe that the attrition count is higher in Male than in Female.
Number of Male employee Attrition: 150
Number of Female employee Attrition: 87
```{r}
#Subset data for attrition and gender.
attr_gender <- hr_data %>%
  group_by(Attrition, Gender) %>%
  summarise(emp_count = n(), na.rm = TRUE) %>%
  ggplot(aes(x = Attrition, y = emp_count, fill = Gender)) +
  labs(x = "Attrition" , y = " Number of Employee ") +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("pink", "darkblue")) +
  ggtitle("Employee Attrition with respect to Gender") +
  geom_text(aes(label = emp_count),
            vjust = -0.2,
            position = position_dodge(0.5))
theme_bw()
#Plot bar graph for attrition with respect to gender.
grid.arrange(attr_gender)

```

**Atrrition count with respect to Departments : **
We observe attrition count in each department as:
HR : 12/63   = 19%
R&D: 133/961 = 13.84%
Sale: 92/446 = 20.63%
The attrition count is higher in Research and development department but the attrition rate is much higher in Sales Department.

```{r}
#Subset data for attrition and department to plot graphs.
dept <- hr_data %>%
  group_by(Department) %>%
  tally() %>%
  ggplot(aes(x = Department, y = n, fill = Department)) +
  labs(x = "Department" , y = " count ") +
  ggtitle("     Employee in each Department ") +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = c("blue", "orange", "yellow")) +
  geom_text(aes(label = n), vjust = 0.3, position = position_dodge(0.5))
  
  dept_attr <- hr_data %>%
  group_by(Department, Attrition) %>%
  tally() %>%
  ggplot(aes(x = Department, y = n, fill = Attrition)) +
  labs(x = "Department" , y = " count ") +
  ggtitle("    Attrition in each Department ") +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = c("green", "red")) +
  geom_text(aes(label = n), vjust = -0.1, position = position_dodge(0.5))
  
  #Plot bar graph for attrition with respect to each department.
  grid.arrange(dept, dept_attr)
```

In this section we will analyze factors that could have been a possible influence on individuals leaving the organization.

**Attrition affected by Work Life Balance : **
WorkLifeBalance : 1- 'Bad' 
                  2- 'Good' 
                  3- 'Better'
                  4- 'Best'

Plotted the table to observe the number of people affected by the work life balance.

```{r}
#Subset data for WorkLifeBalance
wrk_life_bal <- hr_data %>%
  group_by(Attrition, WorkLifeBalance) %>%
  summarise(emp_count = n())

#set theme color for the table
bluecol <- ttheme_minimal(
  core = list(
    bg_params = list(fill = blues9[1:4], col = "navyblue"),
    fg_params = list(fontface = 4)
  ),
  colhead = list(fg_params = list(col = "Red", fontface = 4L)),
  rowhead = list(fg_params = list(col = "NavyBlue", fontface = 2L))
)

#The two tables for employee work life balance with and without attrition.
No_Attr_table <- tableGrob(wrk_life_bal[1:4, 1:3], theme = bluecol)
Yes_Attr_table <- tableGrob(wrk_life_bal[5:8, 1:3], theme = bluecol)

#Plot the table
grid.arrange(top = "Work Life Balance Affecting Attrition ",
             No_Attr_table,
             Yes_Attr_table,
             nrow = 1)
```

**Attrition affected by Job Satisfaction level : **
Job Satisfaction: 1 'Low' 
                  2 'Medium'
                  3 'High'
                  4 'Very High'      
Plotted the table to observe the number of people affected by the job satisfaction level.
```{r}
#Subset data for WorkLifeBalance
job_satis <- hr_data %>%
  group_by(Attrition, JobSatisfaction) %>%
  summarise(emp_count = n())

#set theme color for the table
bluecol <- ttheme_minimal(
  core = list(
    bg_params = list(fill = blues9[1:4], col = "navyblue"),
    fg_params = list(fontface = 4)
  ),
  colhead = list(fg_params = list(col = "Red", fontface = 4L)),
  rowhead = list(fg_params = list(col = "NavyBlue", fontface = 2L))
)

#The two tables for employee work life balance with and without attrition.
job_no_attr_table <- tableGrob(job_satis[1:4, 1:3], theme = bluecol)
job_yes_attr_table <- tableGrob(job_satis[5:8, 1:3], theme = bluecol)

#Plot the table
grid.arrange(top = "Job Satisfaction Affecting Attrition ",
             job_no_attr_table,
             job_yes_attr_table,
             nrow = 1)
```

**Attrition with respect to work life balance and Distance from Home : **
Then analysed the box plot which tells us the median of the data indicating how work life balance and jdistance from home together affects the attrition count in employees.
The attrition mean is higher in employees staying away from home and having bad work life balance.

```{r}
#Subset data for WorkLifeBalance and distancefromhome to plot boxgraph
wrk_life_bal_plot <- ggplot(
  hr_data,
  aes(
    x = WorkLifeBalance,
    y = DistanceFromHome,
    group = WorkLifeBalance,
    fill = WorkLifeBalance
  )
) +
  geom_boxplot(alpha = 0.5, col = "Red") +
  theme(legend.position = "none") +
  theme_classic() +
  facet_wrap( ~ Attrition) +
  ggtitle("Attrition") +
  theme(plot.title = element_text(hjust = 0.5))
wrk_life_bal_plot

```

**Attrition affected by Income : **
Analyse attrition with respect to various income group and observe how attrition is affecting employees at different income level.
```{r}
#Monthly income with respect to Atrrition
income <- ggplot(hr_data,
                 aes(x = MonthlyIncome, fill = Attrition)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("lightgreen", "darkred"))

#Monthly rate income with respect to Atrrition
monthly <- ggplot(hr_data,
                  aes(x = MonthlyRate, fill = Attrition)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("lightgreen", "darkred"))

#Daily rate income with respect to Atrrition
daily <- ggplot(hr_data,
                aes(x = DailyRate, fill = Attrition)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("lightgreen", "darkred"))

#Hourly rate income with respect to Atrrition
hourly <- ggplot(hr_data,
                 aes(x = HourlyRate, fill = Attrition)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("lightgreen", "darkred"))

#Plot each graph in grid format for income, monthly , daily and hourly
grid.arrange(income,
             monthly,
             daily,
             hourly,
             ncol = 2,
             nrow = 2)

```

## Conclusion
The exploratory data analyses conducted on the HR dataset focuses on few factors affection the attrition count on Employee.
The few factors analysed were attrition affected by some internal factors like Gender, different departments in company, income l.evels
External factors affectin g were work life balance , distance from home, job satisfaction.
This Analysis helps HR understand and take necessary steps to reduce the company attrition. 
